---
title: "Supervised Classification using NIR Spectra from Pre-symptomatic and Naive samples"
author: "William Parrott"
date: "2025-05-28"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

*Code adapted from Dr. Anna Conrad*
*REF: https://infometrix.com/pirouette/*

# Setup

```{r}
######Creating a vector of wavelengths#########
library(readr)
library(readxl)

#change to first file in spectra directory; col_names=FALSE if there are no column headers, default is 'TRUE'
wavelength <- read_tsv("C:/Users/willi/Desktop/Master's Thesis Workbook/Summer 2024 Data_Wavelength/wavelengthref.txt", col_names=TRUE) 

#Wavelength vector using column 1
#wavelength <- a[,1] -> not in use, wavelength above generates same vector

#Import data table
df <- read.csv("C:/Users/willi/Desktop/Master's Thesis Workbook/RerunsMay25/Data/August/PhloemAugust2024WLTwoClass.csv")

#Plot all spectra 
library(mdatools)
spectra = df[,4:260]
wavelength = wavelength

par(mfrow = c(1, 1))
attr(spectra, "xaxis.name") = expression("Wavelength, nm")
attr(spectra, "name") = "Fraser Fir Spectra"
attr(spectra, "xaxis.values") = wavelength
attr(spectra,"yaxis.name") = "Reflectance"
mdaplot(spectra, type = "l", xticks = c(2550, 2341, 2043, 1854, 1697, 1564, 1353), xticklabels = c(2550, 2341, 2043, 1854, 1697, 1564, 1353))

```

# Outlier trimming with fda/fda.usc

```{r}
#Code from Heim et al. 2018 and Febrero-Bande and de la Fuente 2012
library(fda)
library(fda.usc)

#data matrix for outlier ID
dat <- as.matrix(df[,4:260]) #raw data

#create list with variable names
dat.label <- list(main="With", xlab="Wavelength, nm", ylab="Reflectance") #label names to have plot information in fdata object

#specify data
#argvals, by default, 1:m where m are the points observed in each curve
myfdata <- fdata(dat, names=dat.label) 

##run outlier detection based on depth measures 
#based on assumption that depth and outlyingness are inverse notions, so if an outlier is in the dataset then the corresponding curve will have a significantly lower depth
set.seed(3456) #set seed for consistent results
out <- outliers.depth.trim(myfdata, dfunc = depth.FM, nb = 10, smo = 0.1, trim = 0.06)

#view the sample rows selected as outliers
FF.out <- out$outliers
FF.out

#remove outliers from full data set
FF_wo.out <- df[-c(57),]

#visualize data with and without outliers; not essential code
dat.label.out <- list(main="Out", xlab="Wavelength, nm", ylab="Reflectance")
#myfdata.out <- myfdata[-c(as.integer(out$outliers)),] #create a second fdata after outlier detection for visualization
myfdata.out <- fdata(FF_wo.out[,4:260], names=dat.label.out)
par(mfrow=c(1,2))
plot(myfdata)
plot(myfdata.out)
```

#Additional Outlier Trimming with Visual Review
```{r}

#Filter additional outliers based on visual review 
boxplot(FF_wo.out$X1442.829394)
boxplot(FF_wo.out$X1696.074188)
#Remove additional outliers if necessary
library(dplyr)
FF_wo.out <- FF_wo.out %>% filter(FF_wo.out$X1442.829394 < 27)
FF_wo.out <- FF_wo.out %>% filter(FF_wo.out$X1696.074188 > 22.5)
#FF_wo.out <- FF_wo.out %>% filter(X1843.743342 < 60)
# First run- skipped the extra outliers because no extremes were present

#Re-plot filtered spectra
spectra = FF_wo.out[,4:260]
wavelength = wavelength

par(mfrow = c(1, 1))
attr(spectra, "xaxis.name") = expression("Wavelength (nm)")
attr(spectra, "name") = "Trimmed Fraser Fir Spectra"
attr(spectra, "xaxis.values") = wavelength
attr(spectra,"yaxis.name") = "Reflectance"
mdaplot(spectra, type = "l", xticks = c(2550, 2341, 2043, 1854, 1697, 1564, 1353), xticklabels = c(2550, 2341, 2043, 1854, 1697, 1564, 1353))

###Group and average 
library(dplyr)
FF_wo.out$Genotype <- factor(FF_wo.out$Genotype)
FF_tol <- FF_wo.out %>% filter(Genotype == "X97")
FF_int <- FF_wo.out %>% filter(Genotype == "X121")

Tolerant <- colMeans(FF_tol[,4:260])
Intolerant <- colMeans(FF_int[,4:260])

dat_avg <- rbind(Tolerant, Intolerant)

spectra = dat_avg
wavelength = wavelength

par(mfrow = c(1, 1))
attr(spectra, "xaxis.name") = expression("Wavelength (nm)")
attr(spectra, "name") = "Average August Phloem Trimmed Fraser Fir Spectra"
attr(spectra, "xaxis.values") = wavelength
attr(spectra,"yaxis.name") = "Transformed Reflectance"
mdaplotg(spectra, type = "l", xticks =  c(2550, 2341, 2043, 1854, 1697, 1564, 1353), xticklabels = c(2550, 2341, 2043, 1854, 1697, 1564, 1353))

#transform spectra to the second derivative
library(mdatools)
FF_trim_2derv_x <- prep.savgol(as.matrix(FF_wo.out[,4:260]), width=15, porder=2, dorder=2)
FF_trim_y <- FF_wo.out[,1:3]
FF_trim_2derv <- as.data.frame(cbind(FF_trim_y, FF_trim_2derv_x))


###Group and plot spectra by symptom type
library(dplyr)

FF_trim_2derv$Genotype <- factor(FF_trim_2derv$Genotype)
FF_tolerant_2derv <- FF_trim_2derv %>% filter(Genotype == "X97")
FF_intolerant_2derv<-FF_trim_2derv %>% filter(Genotype == "X121")

#Average spectra by symptom type
High_Resistance <- colMeans(FF_tolerant_2derv[,4:260])
No_Resistance <- colMeans(FF_intolerant_2derv[,4:260])

##Combine average data into one data frame
dat_2derv <- rbind(High_Resistance, No_Resistance)

spectra = dat_2derv
wavelength = wavelength

#Plot average transformed spectra 
par(mfrow = c(1, 1))
attr(spectra, "xaxis.name") = expression("Wavelength (nm)")
attr(spectra, "name") = "Transformed Average Highly and Not Resistant Phloem Spectra - August"
attr(spectra, "xaxis.values") = wavelength
attr(spectra,"yaxis.name") = "Transformed Reflectance"
mdaplotg(spectra, type = "l", xticks =  c(2550, 2341, 2043, 1854, 1697, 1564, 1353), xticklabels = c(2550, 2341, 2043, 1854, 1697, 1564, 1353))

#Plot differences between average transformed spectra
sub_spec = abs(Tolerant-Intolerant)
diff = data.frame(unlist(wavelength[1]),unlist(sub_spec))
names(diff) = c("Wavelength", "Difference")
ggplot(diff, aes(Wavelength, Difference)) + geom_line()

```

```{r}
#####PCA######
library(ggfortify)
library(dplyr)
library(ggplot2)

FF_pca_2derv <- prcomp(FF_trim_2derv[,4:260])

#PCA summary statistics w/o centering or scaling
summary(FF_pca_2derv)
FF_pca_2derv
screeplot(FF_pca_2derv, npcs=min(10, length(FF_pca_2derv$sdev)), type=c("lines"),
          main=deparse(substitute('Fraser Fir PCA 2nd Derivative')))

#Plot pca w/o frame
#By symptom
autoplot(FF_pca_2derv, data=FF_trim_2derv, colour='Genotype', size=2) + theme_bw()

#Plot pca w/ frame and w/o loadings
autoplot(FF_pca_2derv, data=FF_trim_2derv, colour='Genotype', loadings=FALSE, loadings.colour='black',
         loadings.label=FALSE, loadings.label.size=2, loadings.label.colour='black',
         frame=TRUE, frame.type='norm') + theme_bw()


```

# Support Vector Machine

```{r}
library(caret)
library(e1071)
library(MLmetrics)

#Used for consistency
set.seed(3456)
#p=0.7 -> partition data into 70% training and 30% testing
trainIndex_transformed <- createDataPartition(FF_trim_2derv$Genotype, p =0.7, 
                                              list = FALSE,
                                              times = 1)
head(trainIndex_transformed)

train_2derv <- FF_trim_2derv[ trainIndex_transformed,]
test_2derv <- FF_trim_2derv[-trainIndex_transformed,]

train_y_2derv <- train_2derv[,1]
test_y_2derv <- test_2derv[,1]
train_x_2derv <- train_2derv[,4:260]
test_x_2derv <- test_2derv[,4:260]

dim(train_x_2derv)
length(train_y_2derv)
table(train_y_2derv)
dim(test_x_2derv)
length(test_y_2derv)
table(test_y_2derv)

dat_ff_2derv <- data.frame(x=train_x_2derv, y=as.factor(train_y_2derv))
dat_ff_2derv.te <- data.frame(x=test_x_2derv, y=as.factor(test_y_2derv))

##Determine optimal parameters
set.seed(3456)
ff_transform_tune.out <- e1071::tune(svm, y~., data=dat_ff_2derv, kernel = "radial", 
                                      ranges = list(cost=c(0.01, 0.1, 1, 10, 100), gamma = 5*10^(-2:2)), scale = TRUE)
summary(ff_transform_tune.out)

##Using a radial kernel
set.seed(3456)
#Cross = 10-fold cross validation
out_ff_2derv <- svm(y~., data=dat_ff_2derv, kernel = "radial", cost = 10, gamma = 0.05, cross = 10, scale = TRUE)
summary(out_ff_2derv)
table(out_ff_2derv$fitted, dat_ff_2derv$y)
Accuracy(out_ff_2derv$fitted, dat_ff_2derv$y)

#Predict with best model and testing data set
ff_pred_transform.te <- predict(out_ff_2derv, newdata=dat_ff_2derv.te)
table(ff_pred_transform.te, dat_ff_2derv.te$y)
Accuracy(ff_pred_transform.te, dat_ff_2derv.te$y)

##Using a linear kernel
set.seed(3456)
out_ff_2derv <- svm(y~., data=dat_ff_2derv, kernel = "linear", cost = 10, cross = 10, scale = TRUE)
summary(out_ff_2derv)
table(out_ff_2derv$fitted, dat_ff_2derv$y)
Accuracy(out_ff_2derv$fitted, dat_ff_2derv$y)

#Predict with best model and testing data set
ff_pred_transform.te <- predict(out_ff_2derv, newdata=dat_ff_2derv.te)
table(ff_pred_transform.te, dat_ff_2derv.te$y)
Accuracy(ff_pred_transform.te, dat_ff_2derv.te$y)

##Linear produces best testing results
set.seed(3456)
#Cross = 10-fold cross validation
out_ff_2derv <- svm(y~., data=dat_ff_2derv, kernel = "linear", cost = 10, cross = 10, scale = TRUE)
summary(out_ff_2derv)
table(out_ff_2derv$fitted, dat_ff_2derv$y)
Accuracy(out_ff_2derv$fitted, dat_ff_2derv$y)

#Predict with best model and testing data set
ff_pred_transform.te <- predict(out_ff_2derv, newdata=dat_ff_2derv.te)
table(ff_pred_transform.te, dat_ff_2derv.te$y)
Accuracy(ff_pred_transform.te, dat_ff_2derv.te$y)

################
###ROC Curves###
################
ordVec = c("X121","X97")
library(ROCR)
rocplot = function(pred, truth, ...){
  predob = prediction(pred, truth,label.ordering=ordVec)
  perf =  performance(predob, "tpr", "fpr")
  plot(perf, ...)
}

fitted=attributes(predict(out_ff_2derv, dat_ff_2derv, decision.values = TRUE))$decision.values
par(mfrow = c(1,1))
rocplot(fitted, train_y_2derv, main = "Training Data")

fitted=attributes(predict(out_ff_2derv, dat_ff_2derv.te, decision.values = TRUE)) $decision.values
par(mfrow = c(1,1))
rocplot(fitted, test_y_2derv, main = "SVM 2nd Derivative Testing Data")
abline(0,1)

```


# Random Forest
*Code modified from https://www.blopig.com/blog/2017/04/a-very-basic-introduction-to-random-forests-using-r/ & https://www.guru99.com/r-random-forest-tutorial.html*

```{r}
library(randomForest)
library(VSURF)

#Creating a control model using default parameters
trControl <- trainControl(method = "cv",
                          number = 10,
                          search = "grid", 
                          classProbs  = TRUE,
                          savePredictions = TRUE)

set.seed(1234)
rf_default_2derv <- train(y~.,
                          data = dat_ff_2derv,
                          method = "rf",
                          metric = "Accuracy",
                          trControl = trControl)
print(rf_default_2derv)

#Determining best mtry = the optimal number of random variables to use in the model
set.seed(1234)
tuneGrid_transform <- expand.grid(.mtry = c(1: 10))

rf_mtry_2derv <- train(y~.,
                       data = dat_ff_2derv,
                       method = "rf",
                       metric = "Accuracy",
                       tuneGrid = tuneGrid_transform,
                       trControl = trControl,
                       importance = TRUE,
                       nodesize = 14,
                       ntree = 300)
print(rf_mtry_2derv)
#Kappa compares the observed accuracy with the expected accuracy

#Determining the optimal number of nodes
#maxnodes: Set the maximum amount of terminal nodes in the forest
#importance=TRUE: Whether independent variables importance in the random forest should be assessed
store_maxnode_transform <- list()
tuneGrid_transform <- expand.grid(.mtry = 10)
for (maxnodes in c(20: 30)) {
  set.seed(1234)
  rf_maxnode_transform <- train(y~.,
                                data = dat_ff_2derv,
                                method = "rf",
                                metric = "Accuracy",
                                tuneGrid = tuneGrid_transform,
                                trControl = trControl,
                                importance = TRUE,
                                nodesize = 14,
                                maxnodes = maxnodes,
                                ntree = 300)
  current_iteration <- toString(maxnodes)
  store_maxnode_transform[[current_iteration]] <- rf_maxnode_transform
}
results_maxnode <- resamples(store_maxnode_transform)
summary(results_maxnode)

#Determining the optimal number of trees
store_maxtrees_transform <- list()
for (ntree in c(250, 300, 350, 400, 450, 500, 550, 600, 800, 1000, 2000)) {
  set.seed(5678)
  rf_maxtrees_transform <- train(y~.,
                                 data = dat_ff_2derv,
                                 method = "rf",
                                 metric = "Accuracy",
                                 tuneGrid = tuneGrid_transform,
                                 trControl = trControl,
                                 importance = TRUE,
                                 nodesize = 14,
                                 maxnodes = 20,
                                 ntree = ntree)
  key <- toString(ntree)
  store_maxtrees_transform[[key]] <- rf_maxtrees_transform
}
results_tree_transform <- resamples(store_maxtrees_transform)
summary(results_tree_transform)

##Train optimal model
FF_rf_transform <- train(y~.,
                          data = dat_ff_2derv,
                          method = "rf",
                          metric = "Accuracy",
                          tuneGrid = tuneGrid_transform,
                          trControl = trControl,
                          importance = TRUE,
                          nodesize = 14,
                          ntree = 400,
                          maxnodes = 20)
prediction_2derv<-predict(FF_rf_transform, dat_ff_2derv.te)
confusionMatrix(prediction_2derv, dat_ff_2derv.te$y)


#Test training model
rf_classifier_2derv = randomForest(y~., data = dat_ff_2derv, ntree=400, mtry = 10, importance = TRUE)
rf_classifier_2derv

#Test Testing model
rf_classifier_2derv = randomForest(y~., data = dat_ff_2derv.te, ntree=400, mtry = 10, importance = TRUE)
rf_classifier_2derv

##ROC Curves##
library(pROC)

result.predicted.prob <- predict(FF_rf_transform, dat_ff_2derv.te, type  =  "prob")
result.roc <- roc(dat_ff_2derv.te$y, result.predicted.prob$X97)

plot(tpr ~ fpr,
     coords(result.roc, "all", ret =c("tpr", "fpr"), transpose = FALSE),
     type="l",
     main=("Random Forest 2nd Derivative Testing Data"),
     xlab =  "False positive rate",
     ylab =  "True positive rate")
abline(0,1)

#Filter variables of importance
varImp(FF_rf_transform)


```

# Variable Reduction - VarImp

```{r}
#SVM
#Used for consistency
set.seed(3456)
#p=0.7 -> partition by 70%
trainIndex_transformed <- createDataPartition(FF_trim_2derv$Genotype, p =0.7, 
                                              list = FALSE,
                                              times = 1)
head(trainIndex_transformed)

train_2derv <- FF_trim_2derv[ trainIndex_transformed,]
test_2derv <- FF_trim_2derv[-trainIndex_transformed,]

train_y_2derv <- train_2derv[,1]
test_y_2derv <- test_2derv[,1]
train_x_2derv <- train_2derv[c(55,54,56,53,158,57,52,62,51,119,60,58,121,163,120,157,155,16,122,59)]
test_x_2derv <- test_2derv[c(55,54,56,53,158,57,52,62,51,119,60,58,121,163,120,157,155,16,122,59)]

dim(train_x_2derv)
length(train_y_2derv)
table(train_y_2derv)
dim(test_x_2derv)
length(test_y_2derv)
table(test_y_2derv)


dat_2derv <- data.frame(x=train_x_2derv, y=as.factor(train_y_2derv))
dat_2derv.te <- data.frame(x=test_x_2derv, y=as.factor(test_y_2derv))

set.seed(3456)
FF_transform_tune.out <- e1071::tune(svm, y~., data=dat_2derv, kernel = "radial", 
                                      ranges = list(cost=c(0.01, 0.1, 1, 10, 100), gamma = 5*10^(-2:2)), scale = TRUE)
summary(FF_transform_tune.out)

#Radial Kernel
set.seed(3456)
#Cross means removing parts of the data and running it 10 times
out_2derv <- svm(y~., data=dat_2derv, kernel = "radial", cost = 100, gamma = 0.05, cross = 10, scale = TRUE)
summary(out_2derv)
table(out_2derv$fitted, dat_2derv$y)
Accuracy(out_2derv$fitted, dat_2derv$y)

#Predict with best model and testing data set
FF_pred_2derv.te <- predict(out_2derv, newdata=dat_2derv.te)
table(FF_pred_2derv.te, dat_2derv.te$y)
Accuracy(FF_pred_2derv.te, dat_2derv.te$y)

#Linear Kernel
set.seed(3456)
#Cross means removing parts of the data and running it 10 times
out_2derv <- svm(y~., data=dat_2derv, kernel = "linear", cost = 100, cross = 10, scale = TRUE)
summary(out_2derv)
table(out_2derv$fitted, dat_2derv$y)
Accuracy(out_2derv$fitted, dat_2derv$y)

#Predict with best model and testing data set
FF_pred_2derv.te <- predict(out_2derv, newdata=dat_2derv.te)
table(FF_pred_2derv.te, dat_2derv.te$y)
Accuracy(FF_pred_2derv.te, dat_2derv.te$y)

#Use Linear Kernel for optimal model
set.seed(3456)
#Cross means removing parts of the data and running it 10 times
out_2derv <- svm(y~., data=dat_2derv, kernel = "linear", cost = 100, cross = 10, scale = TRUE)
summary(out_2derv)
table(out_2derv$fitted, dat_2derv$y)
Accuracy(out_2derv$fitted, dat_2derv$y)

#Predict with best model and testing data set
FF_pred_2derv.te <- predict(out_2derv, newdata=dat_2derv.te)
table(FF_pred_2derv.te, dat_2derv.te$y)
Accuracy(FF_pred_2derv.te, dat_2derv.te$y)

#ROC Curves
#ROC for training data
fitted=attributes(predict(out_2derv, dat_2derv, decision.values = TRUE)) $decision.values
par(mfrow = c(1,1))
rocplot(fitted, train_y_2derv, main = "Training Data")

#ROC for testing data
fitted=attributes(predict(out_2derv, dat_2derv.te, decision.values = TRUE)) $decision.values
par(mfrow = c(1,1))
rocplot(fitted, test_y_2derv, main = "SVM VarImp Testing Data")
abline(0,1)

```


# Variable Reduction - VSURF

```{r}
#Used for consistency
set.seed(3456)
#p=0.7 -> partition data into 70% training and 30% testing
trainIndex_transformed <- createDataPartition(FF_trim_2derv$Genotype, p =0.7, 
                                              list = FALSE,
                                              times = 1)
head(trainIndex_transformed)

train_2derv <- FF_trim_2derv[ trainIndex_transformed,]
test_2derv <- FF_trim_2derv[-trainIndex_transformed,]

train_y_2derv <- train_2derv[,1]
train_x_2derv <- train_2derv[,4:260]


set.seed(1234)
ff_vsurf <- VSURF(x = train_x_2derv, y = train_y_2derv, mtry = 10)
summary(ff_vsurf)
plot(ff_vsurf)
#Interpretation variables
ff_vsurf$varselect.interp
#Prediction variables
ff_vsurf$varselect.pred
```

#Random Forest using Interpretation Variables 
```{r}
set.seed(3456)
trainIndex_transformed <- createDataPartition(FF_trim_2derv$Genotype, p =0.7, 
                                              list = FALSE,
                                              times = 1)
head(trainIndex_transformed)

train_2derv <- FF_trim_2derv[ trainIndex_transformed,]
test_2derv <- FF_trim_2derv[-trainIndex_transformed,]

train_y_2derv <- train_2derv[,1]
test_y_2derv <- test_2derv[,1]
train_x_2derv <- train_2derv[c(54)]
test_x_2derv <- test_2derv[c(54)]

dim(train_x_2derv)
length(train_y_2derv)
table(train_y_2derv)
dim(test_x_2derv)
length(test_y_2derv)
table(test_y_2derv)

dat_2derv <- data.frame(x=train_x_2derv, y=as.factor(train_y_2derv))
dat_2derv.te <- data.frame(x=test_x_2derv, y=as.factor(test_y_2derv))

#Using default parameters
trControl <- trainControl(method = "cv",
                          number = 10,
                          search = "grid", 
                          classProbs  = TRUE,
                          savePredictions = TRUE)

set.seed(1234)
rf_default_2derv <- train(y~.,
                          data = dat_2derv,
                          method = "rf",
                          metric = "Accuracy",
                          trControl = trControl)
print(rf_default_2derv)

#Determining best mtry = the optimal number of random variables to use in the model
set.seed(1234)
tuneGrid_transform <- expand.grid(.mtry = c(1: 10))
rf_mtry_2derv <- train(y~.,
                       data = dat_ff_2derv,
                       method = "rf",
                       metric = "Accuracy",
                       tuneGrid = tuneGrid_transform,
                       trControl = trControl,
                       importance = TRUE,
                       nodesize = 14,
                       ntree = 300)
print(rf_mtry_2derv)

#Determining the optimal number of nodes
#maxnodes: Set the maximum amount of terminal nodes in the forest
#importance=TRUE: Whether independent variables importance in the random forest should be assessed
store_maxnode_transform <- list()
tuneGrid_transform <- expand.grid(.mtry = 1)
for (maxnodes in c(20: 30)) {
  set.seed(1234)
  rf_maxnode_transform <- train(y~.,
                                data = dat_ff_2derv,
                                method = "rf",
                                metric = "Accuracy",
                                tuneGrid = tuneGrid_transform,
                                trControl = trControl,
                                importance = TRUE,
                                nodesize = 14,
                                maxnodes = maxnodes,
                                ntree = 300)
  current_iteration <- toString(maxnodes)
  store_maxnode_transform[[current_iteration]] <- rf_maxnode_transform
}
results_maxnode <- resamples(store_maxnode_transform)
summary(results_maxnode)

#Determining the optimal number of trees
store_maxtrees_transform <- list()
for (ntree in c(250, 300, 350, 400, 450, 500, 550, 600, 800, 1000, 2000)) {
  set.seed(5678)
  rf_maxtrees_transform <- train(y~.,
                                 data = dat_ff_2derv,
                                 method = "rf",
                                 metric = "Accuracy",
                                 tuneGrid = tuneGrid_transform,
                                 trControl = trControl,
                                 importance = TRUE,
                                 nodesize = 14,
                                 maxnodes = 20,
                                 ntree = ntree)
  key <- toString(ntree)
  store_maxtrees_transform[[key]] <- rf_maxtrees_transform
}
results_tree_transform <- resamples(store_maxtrees_transform)
summary(results_tree_transform)

##Train optimal model
ff_rf_2derv <- train(y~.,
                      data = dat_2derv,
                      method = "rf",
                      metric = "Accuracy",
                      tuneGrid = tuneGrid_transform,
                      trControl = trControl,
                      importance = TRUE,
                      nodesize = 20,
                      ntree = 1000,
                      maxnodes = 20)
prediction_2derv<-predict(ff_rf_2derv, dat_2derv.te)
confusionMatrix(prediction_2derv, dat_2derv.te$y)


#Test training model
rf_classifier_2derv = randomForest(y~., data = dat_ff_2derv, ntree=1000, mtry = 1, importance = TRUE)
rf_classifier_2derv

#Test Testing model
rf_classifier_2derv = randomForest(y~., data = dat_ff_2derv.te, ntree=1000, mtry = 1, importance = TRUE)
rf_classifier_2derv


#ROC Curves
result.predicted.prob <- predict(ff_rf_2derv, dat_2derv.te, type  =  "prob")
result.roc <- roc(dat_2derv.te$y, result.predicted.prob$X97)

plot(tpr ~ fpr, 
     coords(result.roc, "all", ret =c("tpr", "fpr"), transpose = FALSE),
     type="l",
     main=("Rand. Forest VSURF-Interp. Testing Data"),
     xlab =  "False positive rate",
     ylab =  "True positive rate")
abline(0,1)
```

#Random Forest using Prediction Variables
```{r}
set.seed(3456)
trainIndex_transformed <- createDataPartition(FF_trim_2derv$Genotype, p =0.7, 
                                              list = FALSE,
                                              times = 1)
head(trainIndex_transformed)

train_2derv <- FF_trim_2derv[ trainIndex_transformed,]
test_2derv <- FF_trim_2derv[-trainIndex_transformed,]

train_y_2derv <- train_2derv[,1]
test_y_2derv <- test_2derv[,1]
train_x_2derv <- train_2derv[c(54)]
test_x_2derv <- test_2derv[c(54)]

dim(train_x_2derv)
length(train_y_2derv)
table(train_y_2derv)
dim(test_x_2derv)
length(test_y_2derv)
table(test_y_2derv)

dat_2derv <- data.frame(x=train_x_2derv, y=as.factor(train_y_2derv))
dat_2derv.te <- data.frame(x=test_x_2derv, y=as.factor(test_y_2derv))


#Using default parameters
trControl <- trainControl(method = "cv",
                          number = 10,
                          search = "grid", 
                          classProbs  = TRUE,
                          savePredictions = TRUE)

set.seed(1234)
rf_default_2derv <- train(y~.,
                          data = dat_2derv,
                          method = "rf",
                          metric = "Accuracy",
                          trControl = trControl)
print(rf_default_2derv)

#Determining best mtry = the optimal number of random variables to use in the model
set.seed(1234)
tuneGrid_transform <- expand.grid(.mtry = c(1: 10))
rf_mtry_2derv <- train(y~.,
                       data = dat_ff_2derv,
                       method = "rf",
                       metric = "Accuracy",
                       tuneGrid = tuneGrid_transform,
                       trControl = trControl,
                       importance = TRUE,
                       nodesize = 14,
                       ntree = 300)
print(rf_mtry_2derv)

#Determining the optimal number of nodes
#maxnodes: Set the maximum amount of terminal nodes in the forest
#importance=TRUE: Whether independent variables importance in the random forest should be assessed
store_maxnode_transform <- list()
tuneGrid_transform <- expand.grid(.mtry = 7)
for (maxnodes in c(20: 30)) {
  set.seed(1234)
  rf_maxnode_transform <- train(y~.,
                                data = dat_ff_2derv,
                                method = "rf",
                                metric = "Accuracy",
                                tuneGrid = tuneGrid_transform,
                                trControl = trControl,
                                importance = TRUE,
                                nodesize = 14,
                                maxnodes = maxnodes,
                                ntree = 300)
  current_iteration <- toString(maxnodes)
  store_maxnode_transform[[current_iteration]] <- rf_maxnode_transform
}
results_maxnode <- resamples(store_maxnode_transform)
summary(results_maxnode)

#Determining the optimal number of trees
store_maxtrees_transform <- list()
for (ntree in c(250, 300, 350, 400, 450, 500, 550, 600, 800, 1000, 2000)) {
  set.seed(5678)
  rf_maxtrees_transform <- train(y~.,
                                 data = dat_ff_2derv,
                                 method = "rf",
                                 metric = "Accuracy",
                                 tuneGrid = tuneGrid_transform,
                                 trControl = trControl,
                                 importance = TRUE,
                                 nodesize = 14,
                                 maxnodes = 20,
                                 ntree = ntree)
  key <- toString(ntree)
  store_maxtrees_transform[[key]] <- rf_maxtrees_transform
}
results_tree_transform <- resamples(store_maxtrees_transform)
summary(results_tree_transform)


##Train optimal model
ff_rf_2derv <- train(y~.,
                      data = dat_2derv,
                      method = "rf",
                      metric = "Accuracy",
                      tuneGrid = tuneGrid_transform,
                      trControl = trControl,
                      importance = TRUE,
                      nodesize = 20,
                      ntree = 400,
                      maxnodes = 20)
prediction_2derv<-predict(ff_rf_2derv, dat_2derv.te)
confusionMatrix(prediction_2derv, dat_2derv.te$y)


#Test training model
rf_classifier_2derv = randomForest(y~., data = dat_ff_2derv, ntree=400, mtry = 7, importance = TRUE)
rf_classifier_2derv

#Test Testing model
rf_classifier_2derv = randomForest(y~., data = dat_ff_2derv.te, ntree=400, mtry = 7, importance = TRUE)
rf_classifier_2derv

#ROC Curves
result.predicted.prob <- predict(ff_rf_2derv, dat_2derv.te, type  =  "prob")
result.roc <- roc(dat_2derv.te$y, result.predicted.prob$X97)

plot(tpr ~ fpr,
     coords(result.roc, "all", ret =c("tpr", "fpr"), transpose = FALSE),
     type="l",
     main=("Rand. Forest VSURF-Pred. Testing Data"),
     xlab =  "False positive rate",
     ylab =  "True positive rate")
abline(0,1)
```

#Vsurf - Prediction variables

```{r}
##SVM

set.seed(3456)
FF_transform_tune.out <- e1071::tune(svm, y~., data=dat_2derv, kernel = "radial", 
                                      ranges = list(cost=c(0.01, 0.1, 1, 10, 100), gamma = 5*10^(-2:2)), scale = TRUE)
summary(FF_transform_tune.out)

#Radial Kernel
set.seed(3456)
#Cross means removing parts of the data and running it 10 times
out_2derv <- svm(y~., data=dat_2derv, kernel = "radial", cost =100, gamma = 5, cross = 10, scale = TRUE)
summary(out_2derv)
table(out_2derv$fitted, dat_2derv$y)
Accuracy(out_2derv$fitted, dat_2derv$y)

#Predict with best model and testing data set
FF_pred_2derv.te <- predict(out_2derv, newdata=dat_2derv.te)
table(FF_pred_2derv.te, dat_2derv.te$y)
Accuracy(FF_pred_2derv.te, dat_2derv.te$y)

#Linear Kernel
set.seed(3456)
#Cross means removing parts of the data and running it 10 times
out_2derv <- svm(y~., data=dat_2derv, kernel = "linear", cost =100, cross = 10, scale = TRUE)
summary(out_2derv)
table(out_2derv$fitted, dat_2derv$y)
Accuracy(out_2derv$fitted, dat_2derv$y)

#Predict with best model and testing data set
FF_pred_2derv.te <- predict(out_2derv, newdata=dat_2derv.te)
table(FF_pred_2derv.te, dat_2derv.te$y)
Accuracy(FF_pred_2derv.te, dat_2derv.te$y)

#Use Radial Kernel produces best accuracy
set.seed(3456)
#Cross means removing parts of the data and running it 10 times
out_2derv <- svm(y~., data=dat_2derv, kernel = "radial", cost =100, gamma=5, cross = 10, scale = TRUE)
summary(out_2derv)
table(out_2derv$fitted, dat_2derv$y)
Accuracy(out_2derv$fitted, dat_2derv$y)

#Predict with best model and testing data set
FF_pred_2derv.te <- predict(out_2derv, newdata=dat_2derv.te)
table(FF_pred_2derv.te, dat_2derv.te$y)
Accuracy(FF_pred_2derv.te, dat_2derv.te$y)


#ROC Curves
fitted=attributes(predict(out_2derv, dat_2derv, decision.values = TRUE)) $decision.values
par(mfrow = c(1,1))
rocplot(fitted, train_y_2derv, main = "Training Data")

fitted=attributes(predict(out_2derv, dat_2derv.te, decision.values = TRUE)) $decision.values
par(mfrow = c(1,1))
rocplot(fitted, test_y_2derv, main = "SVM VSURF Testing Data")
abline(0,1)

```

# Spectral Resampling

```{r}
library("prospectr")
wav <- FF_trim_2derv[,4:260]
#Reduce spectra by every 5th wavelength
X.bin<- binning(wav, bin.size = 5)
X.bin

#SVM
set.seed(3456)
#p=0.7 -> partition by 70%
trainIndex_transformed <- createDataPartition(FF_trim_2derv$Genotype, p =0.7, 
                                              list = FALSE,
                                              times = 1)
head(trainIndex_transformed)

train_2derv <- FF_trim_2derv[ trainIndex_transformed,]
test_2derv <- FF_trim_2derv[-trainIndex_transformed,]

train_y_2derv <- train_2derv[,1]
test_y_2derv <- test_2derv[,1]
train_x_2derv <- train_2derv[c(6,11, 16,21, 26, 31, 36, 41, 46, 51, 56,61, 66, 71, 76, 81, 86, 91, 96, 101, 106,111, 116,121, 126,131, 136,141, 146,151, 156,161, 166,	171, 176,181, 186,191, 196,201, 206,211, 216,221, 226,231, 236,241, 246,251, 256,260)]
test_x_2derv <- test_2derv[c(6,11, 16,21, 26, 31, 36, 41, 46, 51, 56,61, 66, 71, 76, 81, 86, 91, 96, 101, 106,111, 116,121, 126,131, 136,141, 146,151, 156,161, 166,	171, 176,181, 186,191, 196,201, 206,211, 216,221, 226,231, 236,241, 246,251, 256,260)]

dim(train_x_2derv)
length(train_y_2derv)
table(train_y_2derv)
dim(test_x_2derv)
length(test_y_2derv)
table(test_y_2derv)

dat_2derv <- data.frame(x=train_x_2derv, y=as.factor(train_y_2derv))
dat_2derv.te <- data.frame(x=test_x_2derv, y=as.factor(test_y_2derv))

set.seed(3456)
FF_transform_tune.out <- e1071::tune(svm, y~., data=dat_2derv, kernel = "radial", 
                                      ranges = list(cost=c(0.01, 0.1, 1, 10, 100), gamma = 5*10^(-2:2)), scale = TRUE)
summary(FF_transform_tune.out)

#Radial Kernel
set.seed(3456)
#Cross means removing parts of the data and running it 10 times
out_2derv <- svm(y~., data=dat_2derv, kernel = "radial", cost =10, gamma = 0.05, cross = 10, scale = TRUE)
summary(out_2derv)
table(out_2derv$fitted, dat_2derv$y)
Accuracy(out_2derv$fitted, dat_2derv$y)

#Predict with best model and testing data set
FF_pred_2derv.te <- predict(out_2derv, newdata=dat_2derv.te)
table(FF_pred_2derv.te, dat_2derv.te$y)
Accuracy(FF_pred_2derv.te, dat_2derv.te$y)

#Linear Kernel
set.seed(3456)
#Cross means removing parts of the data and running it 10 times
out_2derv <- svm(y~., data=dat_2derv, kernel = "linear", cost = 10,  cross = 10, scale = TRUE)
summary(out_2derv)
table(out_2derv$fitted, dat_2derv$y)
Accuracy(out_2derv$fitted, dat_2derv$y)

#Predict with best model and testing data set
FF_pred_2derv.te <- predict(out_2derv, newdata=dat_2derv.te)
table(FF_pred_2derv.te, dat_2derv.te$y)
Accuracy(FF_pred_2derv.te, dat_2derv.te$y)

#Radial Kernel
set.seed(3456)
#Cross means removing parts of the data and running it 10 times
out_2derv <- svm(y~., data=dat_2derv, kernel = "linear", cost =10, gamma=0.05, cross = 10, scale = TRUE)
summary(out_2derv)
table(out_2derv$fitted, dat_2derv$y)
Accuracy(out_2derv$fitted, dat_2derv$y)

#Predict with best model and testing data set
FF_pred_2derv.te <- predict(out_2derv, newdata=dat_2derv.te)
table(FF_pred_2derv.te, dat_2derv.te$y)
Accuracy(FF_pred_2derv.te, dat_2derv.te$y)

#ROC Curves
fitted=attributes(predict(out_2derv, dat_2derv, decision.values = TRUE)) $decision.values
par(mfrow = c(1,1))
rocplot(fitted, train_y_2derv, main = "Training Data")

fitted=attributes(predict(out_2derv, dat_2derv.te, decision.values = TRUE)) $decision.values
par(mfrow = c(1,1))
rocplot(fitted, test_y_2derv, main = "SVM Binning Testing Data")
abline(0,1)
```

# Sparse Partial least squares discriminate analysis


```{r}
##installing biocmanager and mix0mics
if (!requireNamespace("BiocManager", quietly = TRUE))    
install.packages("BiocManager")
## then install mixOmics
BiocManager::install("mixOmics")

library("mixOmics")

set.seed(3456)
#p=0.7 -> partition by 70%
trainIndex_transformed <- createDataPartition(FF_trim_2derv$Genotype, p =0.7, 
                                              list = FALSE,
                                              times = 1)
head(trainIndex_transformed)

train_2derv <- FF_trim_2derv[ trainIndex_transformed,]
test_2derv <- FF_trim_2derv[-trainIndex_transformed,]

train_y_2derv <- train_2derv[,1]
test_y_2derv <- test_2derv[,1]
train_x_2derv <- train_2derv[,4:260]
test_x_2derv <- test_2derv[,4:260]

dim(train_x_2derv)
length(train_y_2derv)
table(train_y_2derv)
dim(test_x_2derv)
length(test_y_2derv)
table(test_y_2derv)

dat_2derv <- data.frame(x=train_x_2derv, y=as.factor(train_y_2derv))

#Using default parameters
results_2derv <- mixOmics::plsda(train_x_2derv, train_y_2derv, ncomp = 10)
#Shows the relationship and similarities between samples
plotIndiv(results_2derv)
#plots the correlations between each variable and the selected components
plotVar(results_2derv)

#Tuning the best model
set.seed(30)
train.plsda_2derv <- perf(results_2derv, validation = "Mfold", folds = 5, 
                          progressBar = FALSE, nrepeat = 50)

#Plot components x error rate 
plot(train.plsda_2derv, col = color.mixo(5:7), sd = TRUE,
     legend.position = "horizontal")

list.keepX <- c(5:10, seq(20, 100, 10))
set.seed(30)
tune.splsda.srbct <- tune.splsda(train_x_2derv, train_y_2derv, ncomp = 5,
                                 validation = 'Mfold',
                                 folds = 5, dist = 'max.dist', progressBar = FALSE,
                                 measure = "BER", test.keepX = list.keepX,
                                 nrepeat = 50)

error <- tune.splsda.srbct$error.rate
error
#Number of components
ncomp <- tune.splsda.srbct$choice.ncomp$ncomp
ncomp

#Number of variables to keep in model
select.keepX <- tune.splsda.srbct$choice.keepX[1:ncomp]
select.keepX

#Run sPLSDA on training set
splsda.final_2derv <- mixOmics::splsda(train_x_2derv, train_y_2derv, ncomp = 3, keepX = select.keepX)
plotIndiv(splsda.final_2derv, ind.names = FALSE, legend = TRUE,
          ellipse = TRUE, title = "Training 2nd Derivative sPLS-DA")

set.seed(30)
perf.final_2derv <- perf(splsda.final_2derv, validation = "Mfold", folds = 5,
                         dist = 'max.dist', nrepeat = 50,
                         progressBar = FALSE)
perf.final_2derv$error.rate
perf.final_2derv$error.rate.class

#Predicting the accuracy using the testing set 
test.predict_2derv <- predict(splsda.final_2derv, test_x_2derv, method = "max.dist")
Prediction_2derv <- test.predict_2derv$class$max.dist[,1]
confusion.mat_2derv = get.confusion_matrix(truth = test_y_2derv, predicted = Prediction_2derv)
get.BER(confusion.mat_2derv)

#Get variables used for the model 
#Shows loading values
ind.match_2derv = match(selectVar(splsda.final_2derv, comp = 3)$name,
                        names(perf.final_2derv$features$stable[[1]]))
Freq = as.numeric(perf.final_2derv$features$stable[[1]][ind.match_2derv])
data.frame(selectVar(splsda.final_2derv, comp = 3)$value, Freq)

ind.match_2derv = match(selectVar(splsda.final_2derv, comp = 2)$name,
                        names(perf.final_2derv$features$stable[[1]]))
Freq = as.numeric(perf.final_2derv$features$stable[[1]][ind.match_2derv])
data.frame(selectVar(splsda.final_2derv, comp = 2)$value, Freq)

ind.match_2derv = match(selectVar(splsda.final_2derv, comp = 1)$name,
                        names(perf.final_2derv$features$stable[[1]]))
Freq = as.numeric(perf.final_2derv$features$stable[[1]][ind.match_2derv])
data.frame(selectVar(splsda.final_2derv, comp = 1)$value, Freq)
```